Omitted lit

### 2022-03-04 mr-review.Rmd
fixed title & organized text chunks

---
title: "Mendelian randomization for causal inference in genome-wide association studies"
author: "Frederick J. Boehm & Xiang Zhou"
date: "`r lubridate::now()`"
bibliography: exported.bib
csl: elsevier-with-titles.csl
link-citations: yes
linkcolor: blue
always_allow_html: true
output: 
  - bookdown::pdf_document2
  - bookdown::html_document2
  - bookdown::word_document2
---

@@

Below, we discuss a number of MR methods that use multiple 
instruments. The overall trend appears to be in the direction of increasingly sophisticated 
modeling of horizontal pleiotropy (and correlated horizontal pleiotropy) while
attempting to maintain scalability and computational efficiency.

@@

One approach to integrating multiple SNP instruments into a MR framework is
through calculation of allele scores [@burgess_use_2013]. @burgess_use_2013 
calculate an unweighted score as the number of exposure-increasing alleles in 
the subject's genotypes. They also calculate a weighted allele score by using 
the exposure effects as weights. For example, a subject with $g_k$ copies of 
exposure-increasing alleles for SNP $k$ has an unweighted allele score $z = \sum_{k = 1}^K g_{k}$ 
and a weighted allele score $z = \sum_{k = 1}^K w_kg_k$.

@burgess_use_2013 found that when weights are obtained from external data or from 
cross-validation or jackknife approaches applied to the analysis data, the allele score 
functions as a single instrumental variable and greatly diminishes the bias compared to that of the 
two-stage least squares estimator [@burgess_review_2017;@angrist1999jackknife].

@@

@@

@bowden_mendelian_2015 adapted Egger regression, originally developed 
to detect bias in meta-analyses, to detect bias from horizontal 
pleiotropy in two-sample MR studies with GWAS summary data 
[@egger1997bias]. The slope coefficient estimates the causal effect of the exposure on the outcome. 
Under the assumption that the SNP-exposure relationship is independent 
of the SNP's pleiotropic effects, Egger's test offers a valid test of 
the null causal hypothesis and consistently estimates the causal 
effect, even if all SNPs are invalid instruments [@bowden_mendelian_2015].

@@

@hemani_automating_2017

@berzuini_bayesian_2020 proposed a Bayesian approach to MR where SNP instruments
are permitted to exhibit pleiotropic effects on the outcome. Working in the setting of 
one-sample MR with individual-level data, they construct their 
model with a shrinkage prior, so that many SNP instruments have no horizontal pleiotropic effects. Their 
method relaxes the exclusion restriction assumption to permit horizontal pleiotropy (for some SNP 
instruments) and permits correlated SNP instruments, unlike many frequentist methods. Posterior inference 
proceeds by Markov chain monte carlo sampling. They also elaborate their model for 
a univariate exposure into that for a bivariate exposure to analyze the causal effects of body mass index and serum phenylalanine levels on blood pressure.

@@

@howey_bayesian_2020 use anchored Bayesian networks to

@@

@jiang_hierarchical_2021, building on research from @newcombe2016jam, present a two-stage hierarchical Bayesian 
model that aims to unify MR and transcriptome-wide association studies frameworks.

@@

## Multivariate exposures in MR analysis

@grant_pleiotropy_2021: we introduce three methods which build on common approaches in the univariable setting: {MVMR}-Robust; {MVMR}-Median; and {MVMR}-Lasso

@@

@mounier_bias_2021
@hartwig_two-sample_2016
@lawlor_commentary_2016
@burgess_bias_2016

@@

@darrous_simultaneous_2021

@@

\begin{eqnarray}
U = X\beta_U \odot Z_U + \epsilon_U\nonumber \\
A = U\nu + B\delta + X\beta_A \odot Z_A + \epsilon_A \nonumber \\
B = U\eta + A\gamma + X\beta_B \odot Z_B + \epsilon_B 
  (\#eq:wwer)
\end{eqnarray}

To develop MR methods that are robust to correlated horizontal pleiotropy, 
@brown_welch-weighted_2021 present Welch-weighted Egger regression. Their flexible model 
accommodates both uni-directional and bi-directional effects between exposure and outcome. 
@brown_welch-weighted_2021 specifies their model via Equations \@ref(eq:wwer). They denote 
the two traits by $A$ and $B$, the SNP genotypes matrix by $X$, and the unobserved confounder by $U$. $Z_U$, $Z_A$ and $Z_B$ denote the indicator vectors of zeros and ones to indicate which . Note that the model is agnostic with regards to which is the 
outcome and which the exposure, so we follow the authors' notation here. The symbol $\odot$ denotes the Hadamard elementwise product.

In Equations \@ref(eq:wwer),

>>>>

---
title: "Mendelian randomization with genome-wide association studies data"
author: "Frederick J. Boehm & Xiang Zhou"
date: "`r lubridate::now()`"
bibliography: exported.bib
csl: elsevier-with-titles.csl
link-citations: yes
linkcolor: blue
always_allow_html: true
output: 
  - bookdown::pdf_document2
  - bookdown::html_document2
  - bookdown::word_document2
---

@@

Below, we discuss a number of MR methods that use multiple 
SNP instruments. The overall trend appears to be in the direction of increasingly sophisticated 
modeling of horizontal pleiotropy (and correlated horizontal pleiotropy) while
attempting to maintain scalability and computational efficiency.

@@

### Methods with Individual-level data

One approach to integrating multiple SNP instruments into a MR framework is
through calculation of allele scores [@burgess_use_2013]. @burgess_use_2013 
calculate an unweighted score as the number of exposure-increasing alleles in 
the subject's genotypes. They also calculate a weighted allele score by using 
the exposure effects as weights. For example, a subject with $g_k$ copies of 
exposure-increasing alleles for SNP $k$ has an unweighted allele score $z = \sum_{k = 1}^K g_{k}$ 
and a weighted allele score $z = \sum_{k = 1}^K w_kg_k$. 
@burgess_use_2013 found that when weights are obtained from external data or from 
cross-validation or jackknife approaches applied to the analysis data, the allele score 
functions as a single instrumental variable and greatly diminishes the bias compared to that of the 
two-stage least squares estimator [@burgess_review_2017;@angrist1999jackknife].

Finally, before considering methods for MR with GWAS summary statistics, we consider 
a new method for one-sample MR. @berzuini_bayesian_2020 proposed a Bayesian approach to MR where SNP instruments
are permitted to exhibit pleiotropic effects on the outcome. Working in the setting of 
one-sample MR with individual-level data, they construct their 
model with a shrinkage prior, so that many SNP instruments have no horizontal pleiotropic effects. Their 
method relaxes the exclusion restriction assumption to permit horizontal pleiotropy (for some SNP 
instruments) and permits correlated SNP instruments, unlike many frequentist methods. Posterior inference 
proceeds by Markov chain monte carlo sampling. They also elaborate their model for 
a univariate exposure into that for a bivariate exposure to analyze the causal effects of body mass index and serum phenylalanine levels on blood pressure.

### Methods with Summary data

The proliferation of publicly available GWAS data coincides with the acceleration
of methods for two-sample MR with GWAS summary statistics. Our timeline highlights some
of the many two-sample MR methods with GWAS summary statistics, with an increasing 
density of methods over time in the last three years (Figure \@ref(fig:timeline)).

```{r, timeline, fig.cap = "Two-sample MR with GWAS summary statistics methods publications density increases over time."}
knitr::include_graphics("figures/timeline.pdf")
```

@@

@bowden_mendelian_2015 adapted Egger regression, originally developed 
to detect bias in meta-analyses, to detect bias from horizontal 
pleiotropy in two-sample MR studies with GWAS summary statistics or individual-level data
[@egger1997bias]. Their approach is termed MR-Egger. They formulate their model as

\begin{eqnarray}
X_i = \sum_{j=1}^J\gamma_jG_{ij} + U_i + \epsilon_{Xi}\nonumber \\
Y_i = \sum_{j=1}^J\alpha_jG_{ij} + \beta X_i + U_i + \epsilon_{Yi}
(\#eq:egger)
\end{eqnarray}

where $U$ designates confounders, $X$ and $Y$ denote the exposure and outcome, respectively, $G$ is a SNP 
instrument, and $\epsilon_X$ and $\epsilon_Y$ are the zero-mean, normally distributed random errors.

@bowden_mendelian_2015 then write the outcome in terms of the SNP instrument $j$ and 
the effects defined in Equation \@ref(eq:egger) and define the new parameter $\Gamma_j = \alpha_j + \beta\gamma_j$ (Equation \@ref(eq:egger1)).

\begin{eqnarray}
Y_i = (\alpha_j + \beta\gamma_j)G_{ij} +  \epsilon'_{Yij} \nonumber\\
= \Gamma_jG_{ij} + \epsilon'_{Yij}
(\#eq:egger1)
\end{eqnarray}

In Equation \@ref(eq:egger1), $\alpha_j = 0$ when SNP $j$ is a valid instrument.

MR-Egger fits the regression model in Equation \@ref(eq:egger2).

\begin{equation}
\hat\Gamma_j = \beta_{0E} + \beta_E \hat\gamma_j
(\#eq:egger2)
\end{equation}

The notation in Equation \@ref(eq:egger2) could be a little misleading. The independent
and dependent variables are $\hat\gamma_j$ and $\hat\Gamma_j$, respectively, while 
the regression parameters are $\beta_{0E}$ and $\beta_E$.

The slope coefficient $\hat\beta_E$ estimates the causal effect of the exposure on the outcome. 
Under the assumption that the SNP-exposure relationship is independent 
of the SNP's pleiotropic effects, Egger's test offers a valid test of 
the null causal hypothesis and consistently estimates the causal 
effect, even if all SNPs are invalid instruments [@bowden_mendelian_2015].

@@

@@

@@

@@

@@

@@

@@

@@

### 2022-03-04 omitted.Rmd
fixed title & organized text chunks

>>>>

@hemani_automating_2017

@howey_bayesian_2020 use anchored Bayesian networks to

@jiang_hierarchical_2021, building on research from @newcombe2016jam, present a two-stage hierarchical Bayesian 
model that aims to unify MR and transcriptome-wide association studies frameworks.

@mounier_bias_2021
@hartwig_two-sample_2016
@lawlor_commentary_2016
@burgess_bias_2016

@darrous_simultaneous_2021

\begin{eqnarray}
U = X\beta_U \odot Z_U + \epsilon_U\nonumber \\
A = U\nu + B\delta + X\beta_A \odot Z_A + \epsilon_A \nonumber \\
B = U\eta + A\gamma + X\beta_B \odot Z_B + \epsilon_B 
  (\#eq:wwer)
\end{eqnarray}

To develop MR methods that are robust to correlated horizontal pleiotropy, 
@brown_welch-weighted_2021 present Welch-weighted Egger regression. Their flexible model 
accommodates both uni-directional and bi-directional effects between exposure and outcome. 
@brown_welch-weighted_2021 specifies their model via Equations \@ref(eq:wwer). They denote 
the two traits by $A$ and $B$, the SNP genotypes matrix by $X$, and the unobserved confounder by $U$. $Z_U$, $Z_A$ and $Z_B$ denote the indicator vectors of zeros and ones to designate which . Note that the model is agnostic with regards to which is the 
outcome and which the exposure, so we follow the authors' notation here. The symbol $\odot$ denotes the Hadamard elementwise product.

In Equations \@ref(eq:wwer),

## Multivariate exposures in MR analysis

@grant_pleiotropy_2021: we introduce three methods which build on common approaches in the univariable setting: {MVMR}-Robust; {MVMR}-Median; and {MVMR}-Lasso
